---
title: "Traffic flow analysis in AUSA toll booths"
author: "Julián Ailán"
params:
  date: "!r Sys.Date()"
output:
  md_document:
    toc: yes
    variant: markdown_github
  pdf_document:
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE, fig.align="center"}
knitr::opts_chunk$set(echo = TRUE)
```

### Objective
This project consists on analyzing the evolution of traffic on AUSA toll booths in Buenos Aires highways. The data being used in this project can be found in the [Buenos Aires Data](https://data.buenosaires.gob.ar/dataset/flujo-vehicular-por-unidades-de-peaje-ausa) site. In particular, the project aims to predict the amount of vehicles going through toll booths for a given date in the future. Concretely, given a week $w_i$ the model will be able to predict the traffic volume on weeks $w_{i+1}$ and $w_{i+2}$.

### Prework
* Unzip the `Flujo-Vehicular-por-Unidades-de-Peaje-AUSA.zip` file provided by the [Buenos Aires Data](https://data.buenosaires.gob.ar/dataset/flujo-vehicular-por-unidades-de-peaje-ausa) site. This .zip file contains eleven .csv files.
* Run the `standardize.values()` function from `utils.R`. This function will perform certain transformations in the unzipped .csv files, and will generate a file called `traffic.csv`. This function takes some minutes to finish (approximatelly .... minutes).

### Structure of the data to use
The information provided consists of eleven files, one for each year from 2008 to 2018 (where the latter only provides information until September 9th). Among these files, there is no strict convention on the columns naming, neither on whether categorical values are stored with uppercase letters or a mixture of uppercase and lowercase. To standardize everything to the same criteria, several transformations were performed. Those transformations are defined in `src/utils.R` and consist of:

* `standardize.columns.and.merge_`: not all of the files have the same column names, nor the same column order. This function standardizes their format, and merges them all in a single .csv file.
* `standardize.values`: among categorical attributes, not all of them have the same value for the same concept, for example you could find the same toll booth name writen in all caps and then all lowers. This function standardizes that, and creates a new dataset.

Great part of the work was devoted to obtaining an homogeneous dataset. The following bullets show the transformation steps followed to transform the eleven heterogeneous and non-standardized files into a single homogeneous one:

* Separators: characters used as column separators were not consistent across the different eleven files, as the vast majority used the ';' character, while the 2018 file used ','.
* Header names: names used for file headers differed among the original eleven files, in aspects such as 
    + Not being consistent when using '_' (underscore) to separate words when the column name consists of more than one word.
    + Not using the same column ordering. 
    + Lastly, not being consistent on whether to use all upper- or all lower-case characters.
* No consistency on how to write categorical values: as time went by, the name for a given toll booth changed. For example, you could find the 'Retiro' toll booth named as either 'RET', 'Retiro' or 'RETIRO', depending on which file you were looking at. The same level of inconsistency can be found for attributes like the day name and vehicle type.
* Time convention: all time values related to hours, minutes and seconds where homogenized to the `HH:MM:SS` format. This action was performed as there were instances in which the format was `H:MM:SS`, or even `HH`.
* Date formats: this particular section was the one that required more time. All eleven files have a column named `FECHA`, but its format was incosistent among groups of files. To get dates right, a combination of reading with different formats and moving them to R's date standard was needed. In the end this column was divided into three new ones, one for each component of a date (year, month, day), to allow the user to perform aggregations in a more flexible way, rather than just by day. What follows is the date format that each file had.
    + Files of 2008 all the way to 2013 used the `MM/DD/YYYY` format.
    + Files of 2014 and 2015 used the `DD/MM/YYYY` format.
    + Files of 2016 and 2017 used the `YYYY-MM-DD`.
    + The file of 2018 also uses `YYYY-MM-DD`, but its rows increase by month and then by date, which ends up leaving the file with an incorrect order. An example of how dates transition on this file is:
```
2018,2018-11-01,JUEVES,21:00:00,22:00:00,DELLEPIANE LINIERS,PESADO,EFECTIVO,28
2018,2018-11-01,JUEVES,21:00:00,22:00:00,DELLEPIANE LINIERS,PESADO,INFRACCION,5
2018,2018-11-01,JUEVES,21:00:00,22:00:00,DELLEPIANE LINIERS,PESADO,AUPASS,59
2018,2018-11-01,JUEVES,21:00:00,22:00:00,DELLEPIANE LINIERS,PESADO,AUPASS,3
...
2018,2018-12-01,VIERNES,23:00:00,00:00:00,RETIRO,PESADO,AUPASS,5
2018,2018-12-01,VIERNES,23:00:00,00:00:00,SALGUERO,LIVIANO,NO COBRADO,10
2018,2018-12-01,VIERNES,23:00:00,00:00:00,SALGUERO,LIVIANO,AUPASS,18
2018,2018-12-01,VIERNES,23:00:00,00:00:00,SARMIENTO,LIVIANO,NO COBRADO,13
2018,2018-12-01,VIERNES,23:00:00,00:00:00,SARMIENTO,LIVIANO,AUPASS,36
2018,2018-01-13,SABADO,00:00:00,01:00:00,ALBERDI,LIVIANO,NO COBRADO,1
2018,2018-01-13,SABADO,00:00:00,01:00:00,ALBERDI,LIVIANO,EFECTIVO,6
2018,2018-01-13,SABADO,00:00:00,01:00:00,ALBERDI,LIVIANO,EXENTO,3
2018,2018-01-13,SABADO,00:00:00,01:00:00,ALBERDI,LIVIANO,EFECTIVO,167
...
2018,2018-01-14,DOMINGO,03:00:00,04:00:00,SALGUERO,LIVIANO,NO COBRADO,10
2018,2018-01-14,DOMINGO,03:00:00,04:00:00,SALGUERO,LIVIANO,AUPASS,9
2018,2018-01-14,DOMINGO,03:00:00,04:00:00,SARMIENTO,LIVIANO,NO COBRADO,14
2018,2018-01-14,DOMINGO,03:00:00,04:00:00,SARMIENTO,LIVIANO,NO COBRADO,1
```

The original .csv files provide in each entry an estimate of the amount of vehicles that went through a certain toll both in an interval of time of one hour. After all the transformations mentioned above the finalized datasets consists of the following columns:

* `Y`: indicates the **year** number.
* `M`: indicates the **month** number.
* `D`: indicates the **day** number.
* `Q`: indicates the **quarter** number.
* `day.name`: indicates the **day of the week**.
* `start.hour`: indicates the **start time of the interval**.
* `end.hour`: indicates the **end time of the interval**. 
* `toll.booth`: indicates the **name of the toll booth**.
* `vehicle.type`: indicates whether the vehicle was a **truck or a car**. 
* `payment.method`: indicates if the **payment method**.
* `amount`: indicates the **number of vehicles** that went through the toll booth at that interval of time, and payed with a given payment method.

In addition to the previously mentioned columns, two new columns were added to include the geographic position of each toll booth via coordinates, `lat` and `long`. The values of each coordinate were not provided by the dataset, and had to be manually searched using Google Maps.

In the end, the first rows from the dataset being used look like this:
```{r}
rm(list = ls()); gc();
# Include the definitions of the methods used throughout the notebook.
source('./src/flow.R')
df <- read.csv('~/Documents/traffic/datasets/traffic.csv')
head(df)
```

### Exploratory analysis

#### Increment in traffic
As a way of starting to know this dataset, we can see there's a difference in the amount of entries for each file in our dataset.
```{r trend}
volume.per.year <- df %>%
    group_by(Y) %>% 
    summarise(total.volume = sum(amount))

p <- ggplot(volume.per.year, 
            aes(volume.per.year$Y, volume.per.year$total.volume)) +
  geom_point(shape = 1) +
  stat_summary(fun.data = mean_cl_normal) +
  labs(x = 'Time', y = 'Number of vehicles') +
  geom_smooth(method = 'lm') +
  theme_bw()

plot(p)
```

From the graph above it's clear there's an order of magnitud of difference between the amount of vehicles in the 2008-2013 period versus the period 2014-2018. This is the result of two factors:

* An increment in the amount of observations for tool booths, in particular those of Avellaneda, Alberdi and Dellepiane. The reason why the amount of observations increased an order of magnitude is not known, and not clarified in the web site that provides the information.

```{bash}
cd ~/Documents/traffic/datasets/flujo-vehicular-por-unidades-de-peaje-ausa
find . -name '*.csv' -exec wc -l {} \;
```


```{r incrementinobservations, fig.width=8}
t <- df[c('toll.booth', 'Y', 'amount')] %>%
  group_by(toll.booth, Y) %>%
  summarise(total.volume = sum(amount))

p <- ggplot(data = t,
            aes(x = Y, y = total.volume, group = toll.booth)) +
  geom_line(aes(linetype = toll.booth, color = toll.booth)) +
  labs(x = NULL, y = NULL, title = 'Amount of vehicles per year') +
  theme_bw()

plot(p)
```

* The addition of new toll booths around 2014 like those of Sarmiento, Retiro and Salguero as illustrated below.

```{r timeframes, fig.width=8}
theme_set(theme_classic())

t <- df[c('toll.booth', 'Y', 'amount')] %>%
  group_by(toll.booth, Y) %>%
  summarise(amount = sum(amount)) %>%
  mutate(
    start_year = min(Y),
    end_year = max(Y)) %>%
  distinct(toll.booth, start_year, end_year)

p <- ggplot(data = t, aes(x = start_year, xend = end_year, 
                          y = toll.booth, group = toll.booth)) +
  geom_dumbbell(color = '#0e668b', colour_x = '#0e668b', size_x = 1.5, 
                colour_xend = '#a3c4dc', size_xend = 0.75) +
  labs(x = NULL, y = NULL, 
       title = 'Addition of new toll booths to the dataset') +
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'),
        plot.background = element_rect(fill='#ffffff'),
        panel.background = element_rect(fill='#ffffff'),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(),
        axis.ticks = element_blank(),
        legend.position = 'top',
        panel.border = element_blank())

plot(p)
```


#### Hourly patterns in toll booths
Traffic flow holds a pattern for each toll booth. Performing an hourly breakdown of traffic for the Alberdi tool booth, shown in the following graph, two traffic peaks can be identified: one around 8am and another one around 6pm. This matches with the times people is commuting to an from work. Another interesting pattern is that traffic keeps relatively still between the time of the two peaks.

```{r breakdownalberdi, fig.width=8}
t <- custom.agg(df[(df$toll.booth == 'ALBERDI'),],
                length(unique(df$Y)),
                'vehicle.type', 'start.hour')

p <- ggplot(data = t, aes(x = format(strptime(t$start.hour,"%H:%M:%S"),'%H'), 
                          y = amount, group = vehicle.type)) +
    labs(title = 'Vehicles per hour in Alberdi toll booth', 
         x = 'start.hour',
         y = 'Number of vehicles') +
    geom_line(aes(linetype = vehicle.type, color = vehicle.type)) +
    theme_bw()

plot(p)
```

The previous graph differenciates between the two different kinds of vehicles: motorbikes/cars and trucks. While both kinds show a similar behavior, the volume of motorbikes/cars is greater than the one of trucks. Although this behavior can be seen in other toll booths like Avellaneda or Sarmiento, there are cases like Retiro, where the amount of trucks is almost the same as the one of cars.

```{r breakdownretiro, fig.width=8}
t <- custom.agg(df[(df$toll.booth == 'RETIRO'),],
                length(unique(df$Y)),
                'vehicle.type', 'start.hour')

p <- ggplot(data = t, aes(x = format(strptime(t$start.hour,"%H:%M:%S"),'%H'), 
                          y = amount, group = vehicle.type)) +
    labs(title = 'Vehicles per hour in Retiro toll booth', 
         x = 'start.hour',
         y = 'Number of vehicles') +
    geom_line(aes(linetype = vehicle.type, color = vehicle.type)) +
    theme_bw()

plot(p)
```

Even though the volume of vehicles going through the Retiro toll booth is approximatelly three times smaller than the one of Alberdi, the ratio Liviano-Pesado (light-heavy) is greater. The reason for this is that Retiro is a a commercial route, located in the road to the Port of Buenos Aires, reason why the volume of heavy vehicles is greater than in other toll booths.

Another interesting aspect to analyze of this dataset is the distribution of drivers commiting infractions (id est, not paying when they cross the toll booth), and whether this is something that only happens for light vehicles or in heavy vehicles too. Below are two charts with the distribution of infractions for both Alberdi and Retiro.
```{r infractionsalberdi, fig.width=8}
t <- custom.agg(df[(df$toll.booth == 'ALBERDI' & 
                      df$payment.method == 'INFRACCION'),],
                length(unique(df$Y)),
                'vehicle.type', 'start.hour')

p <- ggplot(data = t, aes(x = format(strptime(t$start.hour,"%H:%M:%S"),'%H'), 
                          y = amount, group = vehicle.type)) +
    labs(title = 'Infractions per hour in Alberdi toll booth', 
         x = 'start.hour',
         y = 'Number of vehicles') +
    geom_line(aes(linetype = vehicle.type, color = vehicle.type)) +
    theme_bw()

plot(p)
```
```{r infractionsretiro, fig.width=8}
t <- custom.agg(df[(df$toll.booth == 'RETIRO' & 
                      df$payment.method == 'INFRACCION'),],
                length(unique(df$Y)),
                'vehicle.type', 'start.hour')

p <- ggplot(data = t, aes(x = format(strptime(t$start.hour,"%H:%M:%S"),'%H'), 
                          y = amount, group = vehicle.type)) +
    labs(title = 'Infractions per hour in Retiro toll booth', 
         x = 'start.hour',
         y = 'Number of vehicles') +
    geom_line(aes(linetype = vehicle.type, color = vehicle.type)) +
    theme_bw()

plot(p)
```

In both cases, the distribution of infractions seems to follow the same distribution of traffic throughout the day: the moments where there are more infractions correlate with those in which traffic volume is the highest.

#### Evolution of payment methods
Another interesting phenomenom are users moving from paying in cash (EFECTIVO) to using automatic tolls (AUPASS). Looking at the animation below, 2016 starts showing a decline in the number of vehicle owners paying in *cash*. However, since 2014 the use of *automatic tolls* started to steadily increase. As a note, the decline in 2018 on both cash and automatic tolls is due to the fact that 2018 data is still not complete.

```{r paymentsperyearanimated, fig.width=10}
t <- df %>%
  group_by(payment.method, Y) %>%
  filter(payment.method %in% c(
    'EFECTIVO', 'AUPASS', 'NO COBRADO', 'INFRACCION')) %>%
  summarise(total.volume = sum(amount))

a <- ggplot(data = t,
            aes(Y, total.volume, group = payment.method)) +
  geom_line(aes(linetype = payment.method, color = payment.method)) +
  geom_segment(aes(xend = 2018, yend = total.volume), 
               linetype = 2, colour = 'grey') +
  geom_point(size = 2) +
  scale_y_log10() +
  geom_text(aes(x = 2018.2, label = payment.method), hjust = 0) +
  transition_reveal(Y) +
  coord_cartesian(clip = 'off') + 
  labs(title = 'Evolution of top 4 payment methods along the years',
       x = NULL, y = 'Total amount of payments') +
  theme_minimal() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5), 
        legend.position='none')

animate(a)
```

### Predicting next two weeks' traffic volume toll booths


```{r, fig.width=10}
t <- df %>%
  filter(Y == 2016) %>%
  group_by(Y, M, D) %>%
  summarise(total.volume = sum(amount))

# Grouped the amount of vehicles per day in `t`, and now creating a time-series.
# This time series uses 7 as frequency as we're interested in weekly
# seasonalities.
t.ts <- msts(t$total.volume, seasonal.periods = c(7))
fit <- stlf(t.ts)
fc <- forecast(fit)
autoplot(fc)
```

```{r, fig.width=10}
t.ts.components <- ets(t.ts)
autoplot(t.ts.components)
```

### Conclusions and next steps