---
title: "Estimation of time-spent on AUSA toll booths"
author: "Julián Ailán"
params:
  date: "!r Sys.Date()"
output:
  md_document:
    toc: yes
    variant: markdown_github
---

```{r setupandrequiredpackages, include=FALSE, fig.align="center"}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls()); gc();

packages.to.include <- c('dplyr', 'tibble', 'ggplot2', 'ff', 'reshape2')
lapply(packages.to.include, require, character.only = T)

rm(packages.to.include)
```

### Objective
This project consists on estimating time-spent by drivers on AUSA toll booths in Buenos Aires highways. Currently users of these highways experience excessive amount of time waiting to go through toll booths on peak hours. In addition to estimating this metric, an analysis of whether contextual variables like toll-booth fee or oil prices have an impact on the behavior and amount of users commuting through these highways.

### Joining the traffic dataset with the oil dataset
The `traffic.csv` dataset previously analized [here](https://github.com/tulians/traffic/tree/master/descriptive) has a file size of approximately 562MB, while the `oil_prices.csv` dataset is only 11.2KB. We'll be merging both files in an inner join fashion, using the year and month columns via the [`merge()`](https://www.rdocumentation.org/packages/base/versions/3.6.0/topics/merge) function. As a result of this merge each existing row in `traffic.csv` will be repeated 4 times, one per each oil type, thus taking the final, merged, dataset to an approximate size of 2.2GB.

R requires that variables are stored in RAM in its entirety, so managing a dataframe this big could be imposible for some machines if no alternate processing is performed. For this reason is that the [`ff`](https://cran.r-project.org/web/packages/ff/index.html) package will be used.

```{r readingunifiedfile, echo=FALSE}
datasets.dir <- '~/Documents/traffic/datasets'
# Directory where `ff` stores temporary files.
options(fftempdir = file.path(datasets.dir, 'tmp'))
unified.ff <- read.csv.ffdf(file = file.path(datasets.dir, 'unified.csv'))
```

### Analysis of oil price behavior
The oil prices dataset consists of monthly prices of different types of oil: super, premium, gasoil, and euro. The four of them experienced a steady increase since 2008 until 2018, which is illustrated in Figure 1.

```{r pricethroughtime, echo=FALSE}
oil_price <- unified.ff %>% as_tibble() %>%
  group_by(Y, oil.type) %>%
  summarise(price = max(price))

ggplot(data = oil_price, aes(x = Y, y = price, group = oil.type)) +
  geom_line(aes(color = oil.type)) +
  geom_point(alpha = 1/5) +
  labs(x = NULL, y = 'ARS ($)',
       title = 'Price of oil through time', caption = 'Figure 1') +
  theme_minimal()

rm(oil_price)
```

```{r correlations, echo=FALSE}
ff.df <- unified.ff %>% as_tibble() %>%
  select(Y, M, amount, price) %>%
  filter(Y >= 2014) %>%
  group_by(Y, M) %>%
  summarise(
    price = mean(price),
    vehicles = sum(amount)
  )

pv.correlation <- cor(ff.df$price, ff.df$vehicles)
```

Building on top of the previous analysis of traffic patterns performed [here](https://github.com/tulians/traffic/blob/master/descriptive/README.md), it would be intersting to analyze whether there was a correlation between the rate at which the number of vehicles passing through toll booths [started to grow month-over-month](https://github.com/tulians/traffic/blob/master/descriptive/README_files/figure-markdown_github/trend-1.png), and the month-over-month (M-o-M) increase in oil prices depicted in Figure 1. As seen in Figure 2 (condidering only information from January 2014 onwards, after the [unexpected growth](https://github.com/tulians/traffic/tree/master/descriptive#increment-in-traffic)) there is barely a linear relation between these two differences, which is represented by the Pearson correlation coefficient of `r pv.correlation`. This correlation, even though it's not strong, it's positive, which implies that an increase in oil prices does not necessarily result in a decrease in traffic volume, but rather the opposite in this case.

```{r correlationgraph, echo=FALSE}
ggplot(data = ff.df, aes(x = price, y = vehicles)) +
  geom_point(alpha = 2/5) +
  geom_smooth(method = 'lm') +
  labs(x = 'Oil price (ARS)', 
       y = 'Volume of vehicles',
       title = 'Relation of oil price and amount of vehicles', 
       caption = 'Figure 2') +
  theme_minimal()
```

```{r differences, echo=FALSE}
ff.df.differences <- data.frame(
  Y = diff(ff.df$Y),
  M = diff(ff.df$M),
  price = diff(ff.df$price / max(ff.df$price)),
  vehicles = diff(ff.df$vehicles / max(ff.df$vehicles))
)

ff.df.differences$vehicles[which.min(ff.df.differences$vehicles)] <- 
  mean(ff.df.differences$vehicles)
ff.df.differences$vehicles[which.max(ff.df.differences$vehicles)] <- 
  mean(ff.df.differences$vehicles)

# Move from wide to tall format to plot multiple series in the same graph.
ff.df.tall <- melt(
  ff.df.differences, 
  id.vars = c('Y', 'M'),
  measure.vars = c('price', 'vehicles'),
  variable.name = 'series',
  value.name = 'measurement') %>%
  select(series, measurement) %>%
  group_by(series) %>%
  mutate(id = row_number())

diff.correlation <- cor(ff.df.differences$price, ff.df.differences$vehicles)
```

A more interesting approach in the search for a relation between vehicle volume and oil price would be to look at the M-o-M differences of each time series. Figure 3 shows the behavior through time of the difference of each of the variables, which was computed over the normalized series. Again, computing the correlation between both differences yields a positive number, `r diff.correlation`, which shows non-conclusive evidence that an increase in oil price could impact the volume of vehicles passing through toll booths (at least not in a month by month basis).

```{r differencesgraph, echo=FALSE}
ggplot(data = ff.df.tall, 
       aes(x = id, y = measurement, group = series, colour = series)) +
  geom_line() +
  labs(x = 'index', 
       y = 'Delta (M_i - M_{i-1})',
       title = 'M-o-M differences for both amount of vehicles and oil prices', 
       caption = 'Figure 3') +
  theme_minimal()

rm(ff.df.differences, ff.df.tall, pv.correlation, diff.correlation, ff.df)
```

### Time spent waiting on queues estimation
It's usual to see vehicles queuing in Buenos Aires' toll booth plazas at peak hours. More than 1.5M vehicles enter the city each day, most of which are people commuting to work. In this context toll booths regularly feature queues that extend through kilometers in highways. The objective of this section is to accurately estimate the average time a given driver waits in order go through the toll booth.

In order to achieve this goal we need to have concrete figures of both the rate at which cars arrive to the toll booths $\lambda$, and also the rate at which cars can be serviced $\mu$, meaning, the amount of cars that go through the toll booth at a given period of time. Both of the rates would be averages, as they will be considering times where there is a long queue, as well as times where there's no queue at all. The $\rho=\lambda/\mu$ ratio, known as traffic intensity, will help us identify the average behavior of the queue, given that if $\rho < 1$ there is a finite probability that the queue can be handled by the booth; on the other hand if $\rho \geq 1$ the queue length will become longer and longer without limit (at least in theoretical terms).

Given that we have information of traffic flow along several toll booths, we'll use it to derive these two coefficients, and find the value of $\rho$ for each of the toll booth plazas. Depending on the dimension of this ratio, we'll need to compute waiting time one way or another. More details on this calculations will be given in future sections.

#### Important considerations
##### Toll booth plaza naming, geolocation, and amount of toll booths per plaza
Up to now we've been working with the dataset without validating whether all the information provided was accurate. However, there were mentions in the [previous section](https://github.com/tulians/traffic/tree/master/descriptive#increment-in-traffic) to strange patterns in data. In this section we'll validate that every toll booth plaza name provided in the dataset actually exists, and also where it's located geographically. To perform this validation we'll use a [map](https://www.ausa.com.ar/documentos/AUSA-Mapa-Autopistas.pdf) provided by the same company the information is about.

* `Alberdi`: _Juan Bautista Alberdi_ is an avenue located at [(-34.6429211, -58.4910398)](https://www.google.com/maps/@-34.6429211,-58.4910398,18z) which provides a way to enter the _Perito Moreno_ highway. However, there is no toll booth in such entrance. Looking at the previously mentioned [map](https://www.ausa.com.ar/documentos/AUSA-Mapa-Autopistas.pdf), we can validate there is no toll booth sign in the entrance. However, there is such a sign in an actual toll booth in the _25 de Mayo_ highway, located at [(-34.6252954,-58.4022763)](https://www.google.com/maps/@-34.6252954,-58.4022763,17z), under the name of _Peaje Alberti_ (mind that they differ in one character, as the latter has a _t_ instead of a _d_). Given this context, most likely the information we see for _Alberdi_ corresponds to _Alberti_, as the latter is the one that has a toll booth, while the former does not. Something important to mention about this toll booth is that it counts with [3 lanes](https://www.google.com/maps/@-34.6253084,-58.399961,3a,75y,296.16h,87.96t/data=!3m7!1e1!3m5!1szK1wuTFcAlvJddIHqUlZWw!2e0!6s%2F%2Fgeo0.ggpht.com%2Fcbk%3Fpanoid%3DzK1wuTFcAlvJddIHqUlZWw%26output%3Dthumbnail%26cb_client%3Dmaps_sv.tactile.gps%26thumb%3D2%26w%3D203%26h%3D100%26yaw%3D317.5315%26pitch%3D0%26thumbfov%3D100!7i13312!8i6656) to enter the city, and [2 lanes](https://www.google.com/maps/@-34.6257901,-58.4001986,3a,75y,84.02h,99.26t/data=!3m6!1e1!3m4!1sTeolsLKGK9ckp3WVWxFLQg!2e0!7i13312!8i6656) to leave. For the sake of the code, I'll keep on using the `Alberdi` label, but everything will be computed with _Peaje Alberti_ in mind.

* `Avellaneda`: the _Parque Avellaneda_ toll booth is located at [(-34.6483842,-58.4782827)](https://www.google.com/maps/place/Toll+Parque+Avellaneda/@-34.6483842,-58.4782827,15z/data=!4m5!3m4!1s0x95bcc976fa19271d:0x114032996c02ca46!8m2!3d-34.6478475!4d-58.477942) in the _Perito Moreno_ highway. This geolocation matches the location provided in the [map](https://www.ausa.com.ar/documentos/AUSA-Mapa-Autopistas.pdf). This toll booth is much bigger than the one of _Peaje Alberti_, as it's right in the highway, and consists of [16 lanes](https://www.google.com/maps/@-34.6485245,-58.4775302,3a,82.3y,306.63h,84.59t/data=!3m6!1e1!3m4!1s0t2jEnNc2pbxYu3mgSMYiw!2e0!7i13312!8i6656) to enter the city and [17 lanes](https://www.google.com/maps/@-34.6473728,-58.4783124,3a,70.3y,140.7h,90.17t/data=!3m6!1e1!3m4!1s8HtLVelrUWM_-Lq3uKhZJw!2e0!7i13312!8i6656) to leave.

* `Dellepiane`: the _Dellepiane_ toll booth is located at [(-34.6476526,-58.4642902)](https://www.google.com/maps/@-34.6476526,-58.4642902,3a,75y,183.74h,83.19t/data=!3m7!1e1!3m5!1sAJy89f4OeGWzUY4j5jP_kA!2e0!6s%2F%2Fgeo2.ggpht.com%2Fcbk%3Fpanoid%3DAJy89f4OeGWzUY4j5jP_kA%26output%3Dthumbnail%26cb_client%3Dmaps_sv.tactile.gps%26thumb%3D2%26w%3D203%26h%3D100%26yaw%3D278.81256%26pitch%3D0%26thumbfov%3D100!7i13312!8i6656) in the _25 de Mayo_ highway. Just like with `Avellaneda`, it's geolocation matches the locationprovided in the [map](https://www.ausa.com.ar/documentos/AUSA-Mapa-Autopistas.pdf). What's particular about this toll booth is that it only consists of [8 lanes](https://www.google.com/maps/@-34.6476526,-58.4642902,3a,75y,183.74h,83.19t/data=!3m7!1e1!3m5!1sAJy89f4OeGWzUY4j5jP_kA!2e0!6s%2F%2Fgeo2.ggpht.com%2Fcbk%3Fpanoid%3DAJy89f4OeGWzUY4j5jP_kA%26output%3Dthumbnail%26cb_client%3Dmaps_sv.tactile.gps%26thumb%3D2%26w%3D203%26h%3D100%26yaw%3D278.81256%26pitch%3D0%26thumbfov%3D100!7i13312!8i6656) to leave the city, but none to enter, most likely due to its proximity to the Avellaneda toll booth.

* `Illia` and `Retiro`: as per the [map](https://www.ausa.com.ar/documentos/AUSA-Mapa-Autopistas.pdf), the _Retiro_ toll booth is located in the _President Arturo Umberto Illia_ highway. Contrary to what was mentioned about this two toll booths in the the [previous analysis](https://github.com/tulians/traffic/tree/master/descriptive#increment-in-traffic), given that there is no distinction between those two labels in the official map, they will be considered to be the same toll booth in this analysis. This toll booth has [16 lanes](https://www.google.com/maps/@-34.5752154,-58.3939207,3a,75y,97.32h,90.8t/data=!3m6!1e1!3m4!1sgu6cZza2fn1MaGwSQDCN8Q!2e0!7i13312!8i6656) to enter the city, and [13 lanes](https://www.google.com/maps/@-34.5753211,-58.3920502,3a,60y,297.07h,84.14t/data=!3m6!1e1!3m4!1syPX2FmTEXJDiDXbKJ_60lw!2e0!7i13312!8i6656) to leave.

* `Sarmiento` and `Salguero`: these two toll booths are the most recent of all, and are completely automatic, thus don't have any kind of barriers of physical toll booths. They rely on a framework that identifies via laser and RFID whether a given car is suscribed to the automatic toll booth pay a fine. Given the fact that this system was recently implemented, and that Google Maps' most up to date photograph is from [2014](https://www.google.com/maps/@-34.5720142,-58.4003746,3a,75y,86.66h,84.34t/data=!3m6!1e1!3m4!1s4EUI6eAipzhLxajyKJyH3Q!2e0!7i13312!8i6656) there is no information on the amount of lanes the system uses, but judging for the [way it is explained in a local newspaper](https://www.clarin.com/brandstudio/autopistas-barreras-funcionan-beneficios_0_BFyjtPTT7.html) the lasers most likely take the width of the highway, which is [4 lanes](https://www.google.com/maps/@-34.572059,-58.4002728,3a,75y,89.44h,76.27t/data=!3m6!1e1!3m4!1sUBDmnfT7MIlSI-IaUNA8rg!2e0!7i13312!8i6656) to enter the city and [4 lanes](https://www.google.com/maps/@-34.5717317,-58.4002346,3a,75y,318.34h,75.16t/data=!3m7!1e1!3m5!1sB5bOs_1b8Kgeal35aCFRdw!2e0!6s%2F%2Fgeo3.ggpht.com%2Fcbk%3Fpanoid%3DB5bOs_1b8Kgeal35aCFRdw%26output%3Dthumbnail%26cb_client%3Dmaps_sv.tactile.gps%26thumb%3D2%26w%3D203%26h%3D100%26yaw%3D9.781906%26pitch%3D0%26thumbfov%3D100!7i13312!8i6656) to leave.

##### Assumptions
As you can see from the above links the amount of toll booths per lane was manually counted using Google Maps. However, these amounts are not fixed, and depend on date and time, in order to accomodate the service to the incoming volume of vehicles. The are no official communications from AUSA on how those changes are performed, so for the sake of this analysis the average amount of lanes which will be mostly active during peak hours.

Building on this last point, it will be assumed the vast majority of the traffic during the morning peak hour is heading towards the city, while the traffic during the afternoon peak hour is leaving the city.

```{r averagearrivalrate, echo=FALSE}
arrival.rate <- unified.ff %>% as_tibble() %>%
  filter((Y >= 2014)
         & (toll.booth == 'AVELLANEDA')) %>%
  select(Y, M, D, amount) %>%
  group_by(Y, M, D) %>%
  summarise(
    vehicles = sum(amount)
  )

minutes.in.a.day <- 24 * 60
# Most likely I should compute all this for each toll both.
# Amount of cars per minute.
lambda <- mean(arrival.rate$vehicles) / minutes.in.a.day
# Cars served in a minute.
# mu <- yet to define. But can be though in a similar fashion. We have the
# amount of cars that went through the booths in a day, and we have the amount
# of booths that a given toll booth plaza has. Dividing this amount of cars
# per day by the amount of booths you get the amount of cars per booth per day,
# and if you divide that by (24 * 60) you get the amount of cars per minute.
# This way you can divide both coefficients and get rho.
```

