---
title: "Estimation of time-spent on AUSA toll booths"
author: "Julián Ailán"
params:
  date: "!r Sys.Date()"
output:
  github_document:
    toc: yes
    pandoc_args: --webtex
---

```{r setupandrequiredpackages, include=FALSE, fig.align="center"}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls()); gc();

packages_to_include <- c(
  "dplyr", "tibble", "ggplot2", "ff", "reshape2", "tseries"
)

lapply (
  packages_to_include, # From this list ...
  require, # ... include them as packages.
  character.only = T
)

rm(packages_to_include)
```

### Objective
This project consists on estimating time-spent by drivers on AUSA toll booths in Buenos Aires highways. Currently users of these highways experience excessive amount of time waiting to go through toll booths on peak hours. In addition to estimating this metric, an analysis of whether contextual variables like toll-booth fee or oil prices have an impact on the behavior and amount of users commuting through these highways.

### Joining the traffic dataset with the oil dataset
The `traffic.csv` dataset previously analized [here](https://github.com/tulians/traffic/tree/master/descriptive) has a file size of approximately 562MB, while the `oil_prices.csv` dataset is only 11.2KB. We'll be merging both files in an inner join fashion, using the year and month columns via the [`merge()`](https://www.rdocumentation.org/packages/base/versions/3.6.0/topics/merge) function. As a result of this merge each existing row in `traffic.csv` will be repeated 4 times, one per each oil type, thus taking the final, merged, dataset to an approximate size of 2.2GB.

R requires that variables are stored in RAM in its entirety, so managing a dataframe this big could be imposible for some machines if no alternate processing is performed. For this reason is that the [`ff`](https://cran.r-project.org/web/packages/ff/index.html) package will be used.

```{r readingunifiedfile, echo=FALSE}
datasets_dir <- "../datasets"
# Directory where `ff` stores temporary files. Take into account temporary files
# will fill up this directory through time. Make sure to delete them once in a
# while.
options(
  fftempdir = file.path(datasets_dir, "tmp")
)
unified_ff <- read.csv.ffdf (
  file = file.path(datasets_dir, "unified.csv")
)
rm(datasets_dir)
```

### Time spent waiting on queues estimation
It's usual to see vehicles queuing in Buenos Aires' toll booth plazas at peak hours. More than 1.5M vehicles enter the city each day, most of which are people commuting to work. In this context toll booths regularly feature queues that extend through kilometers in highways. The objective of this section is to accurately estimate the average time a given driver waits in order go through the toll booth.

In order to achieve this goal we need to have concrete figures of both the rate at which cars arrive to the toll booths $\lambda$, and also the rate at which cars can be serviced $\mu$, meaning, the amount of cars that go through the toll booth at a given period of time. Both of the rates would be averages, as they will be considering times where there is a long queue, as well as times where there's no queue at all. The $\rho=\lambda/\mu$ ratio, known as traffic intensity, will help us identify the average behavior of the queue, given that if $\rho < 1$ there is a finite probability that the queue can be handled by the booth; on the other hand if $\rho \geq 1$ the queue length will become longer and longer without limit up to infinity (at least in theoretical terms).

Given that we have information of traffic flow along several toll booths, we'll use it to derive these two coefficients, and find the value of $\rho$ for each of the toll booth plazas. Depending on the dimension of this ratio, we'll need to compute waiting time one way or another. More details on this calculations will be given in future sections.

#### Important considerations
##### Toll booth plaza naming, geolocation, and amount of toll booths per plaza
Up to now we've been working with the dataset without validating whether all the information provided was accurate. However, there were mentions in the [previous section](https://github.com/tulians/traffic/tree/master/descriptive#increment-in-traffic) to strange patterns in data. In this section we'll validate that every toll booth plaza name provided in the dataset actually exists, and also where it's located geographically. To perform this validation we'll use a [map](https://www.ausa.com.ar/documentos/AUSA-Mapa-Autopistas.pdf) provided by the same company the information is about.

* `Alberdi`: _Juan Bautista Alberdi_ is an avenue located at [(-34.6429211, -58.4910398)](https://www.google.com/maps/@-34.6429211,-58.4910398,18z) which provides a way to enter the _Perito Moreno_ highway. However, there is no toll booth in such entrance. Looking at the previously mentioned [map](https://www.ausa.com.ar/documentos/AUSA-Mapa-Autopistas.pdf), we can validate there is no toll booth sign in the entrance. However, there is such a sign in an actual toll booth in the _25 de Mayo_ highway, located at [(-34.6252954,-58.4022763)](https://www.google.com/maps/@-34.6252954,-58.4022763,17z), under the name of _Peaje Alberti_ (mind that they differ in one character, as the latter has a _t_ instead of a _d_). Given this context, most likely the information we see for _Alberdi_ corresponds to _Alberti_, as the latter is the one that has a toll booth, while the former does not. Something important to mention about this toll booth is that it counts with [3 lanes](https://www.google.com/maps/@-34.6253084,-58.399961,3a,75y,296.16h,87.96t/data=!3m7!1e1!3m5!1szK1wuTFcAlvJddIHqUlZWw!2e0!6s%2F%2Fgeo0.ggpht.com%2Fcbk%3Fpanoid%3DzK1wuTFcAlvJddIHqUlZWw%26output%3Dthumbnail%26cb_client%3Dmaps_sv.tactile.gps%26thumb%3D2%26w%3D203%26h%3D100%26yaw%3D317.5315%26pitch%3D0%26thumbfov%3D100!7i13312!8i6656) to enter the city, and [2 lanes](https://www.google.com/maps/@-34.6257901,-58.4001986,3a,75y,84.02h,99.26t/data=!3m6!1e1!3m4!1sTeolsLKGK9ckp3WVWxFLQg!2e0!7i13312!8i6656) to leave. For the sake of the code, I'll keep on using the `Alberdi` label, but everything will be computed with _Peaje Alberti_ in mind.

* `Avellaneda`: the _Parque Avellaneda_ toll booth is located at [(-34.6483842,-58.4782827)](https://www.google.com/maps/place/Toll+Parque+Avellaneda/@-34.6483842,-58.4782827,15z/data=!4m5!3m4!1s0x95bcc976fa19271d:0x114032996c02ca46!8m2!3d-34.6478475!4d-58.477942) in the _Perito Moreno_ highway. This geolocation matches the location provided in the [map](https://www.ausa.com.ar/documentos/AUSA-Mapa-Autopistas.pdf). This toll booth is much bigger than the one of _Peaje Alberti_, as it's right in the highway, and consists of [16 lanes](https://www.google.com/maps/@-34.6485245,-58.4775302,3a,82.3y,306.63h,84.59t/data=!3m6!1e1!3m4!1s0t2jEnNc2pbxYu3mgSMYiw!2e0!7i13312!8i6656) to enter the city and [17 lanes](https://www.google.com/maps/@-34.6473728,-58.4783124,3a,70.3y,140.7h,90.17t/data=!3m6!1e1!3m4!1s8HtLVelrUWM_-Lq3uKhZJw!2e0!7i13312!8i6656) to leave.

* `Dellepiane`: the _Dellepiane_ toll booth is located at [(-34.6476526,-58.4642902)](https://www.google.com/maps/@-34.6476526,-58.4642902,3a,75y,183.74h,83.19t/data=!3m7!1e1!3m5!1sAJy89f4OeGWzUY4j5jP_kA!2e0!6s%2F%2Fgeo2.ggpht.com%2Fcbk%3Fpanoid%3DAJy89f4OeGWzUY4j5jP_kA%26output%3Dthumbnail%26cb_client%3Dmaps_sv.tactile.gps%26thumb%3D2%26w%3D203%26h%3D100%26yaw%3D278.81256%26pitch%3D0%26thumbfov%3D100!7i13312!8i6656) in the _25 de Mayo_ highway. Just like with `Avellaneda`, it's geolocation matches the locationprovided in the [map](https://www.ausa.com.ar/documentos/AUSA-Mapa-Autopistas.pdf). What's particular about this toll booth is that it only consists of [8 lanes](https://www.google.com/maps/@-34.6476526,-58.4642902,3a,75y,183.74h,83.19t/data=!3m7!1e1!3m5!1sAJy89f4OeGWzUY4j5jP_kA!2e0!6s%2F%2Fgeo2.ggpht.com%2Fcbk%3Fpanoid%3DAJy89f4OeGWzUY4j5jP_kA%26output%3Dthumbnail%26cb_client%3Dmaps_sv.tactile.gps%26thumb%3D2%26w%3D203%26h%3D100%26yaw%3D278.81256%26pitch%3D0%26thumbfov%3D100!7i13312!8i6656) to leave the city, but none to enter, most likely due to its proximity to the Avellaneda toll booth.

* `Illia` and `Retiro`: as per the [map](https://www.ausa.com.ar/documentos/AUSA-Mapa-Autopistas.pdf), the _Retiro_ toll booth is located in the _President Arturo Umberto Illia_ highway. Contrary to what was mentioned about this two toll booths in the the [previous analysis](https://github.com/tulians/traffic/tree/master/descriptive#increment-in-traffic), given that there is no distinction between those two labels in the official map, they will be considered to be the same toll booth in this analysis. This toll booth has [16 lanes](https://www.google.com/maps/@-34.5752154,-58.3939207,3a,75y,97.32h,90.8t/data=!3m6!1e1!3m4!1sgu6cZza2fn1MaGwSQDCN8Q!2e0!7i13312!8i6656) to enter the city, and [13 lanes](https://www.google.com/maps/@-34.5753211,-58.3920502,3a,60y,297.07h,84.14t/data=!3m6!1e1!3m4!1syPX2FmTEXJDiDXbKJ_60lw!2e0!7i13312!8i6656) to leave.

* `Sarmiento` and `Salguero`: these two toll booths are the most recent of all, and are completely automatic, thus don't have any kind of barriers of physical toll booths. They rely on a framework that identifies via laser and RFID whether a given car is suscribed to the automatic toll booth pay a fine. Given the fact that this system was recently implemented, and that Google Maps' most up to date photograph is from [2014](https://www.google.com/maps/@-34.5720142,-58.4003746,3a,75y,86.66h,84.34t/data=!3m6!1e1!3m4!1s4EUI6eAipzhLxajyKJyH3Q!2e0!7i13312!8i6656) there is no information on the amount of lanes the system uses, but judging for the [way it is explained in a local newspaper](https://www.clarin.com/brandstudio/autopistas-barreras-funcionan-beneficios_0_BFyjtPTT7.html) the lasers most likely take the width of the highway, which is [4 lanes](https://www.google.com/maps/@-34.572059,-58.4002728,3a,75y,89.44h,76.27t/data=!3m6!1e1!3m4!1sUBDmnfT7MIlSI-IaUNA8rg!2e0!7i13312!8i6656) to enter the city and [4 lanes](https://www.google.com/maps/@-34.5717317,-58.4002346,3a,75y,318.34h,75.16t/data=!3m7!1e1!3m5!1sB5bOs_1b8Kgeal35aCFRdw!2e0!6s%2F%2Fgeo3.ggpht.com%2Fcbk%3Fpanoid%3DB5bOs_1b8Kgeal35aCFRdw%26output%3Dthumbnail%26cb_client%3Dmaps_sv.tactile.gps%26thumb%3D2%26w%3D203%26h%3D100%26yaw%3D9.781906%26pitch%3D0%26thumbfov%3D100!7i13312!8i6656) to leave.

##### Assumptions
* As you can see from the above links the amount of toll booths per lane was manually counted using Google Maps. However, these amounts are not fixed, and depend on date and time, in order to accomodate the service to the incoming volume of vehicles. The are no official communications from AUSA on how those changes are performed, so for the sake of this analysis the amount of toll booths to use would be the sum of those going in and those going out of the city. This is due to the fact that the information provided is aggregated to a level that does not provide visibility on the direction of the vehicles.

* Building on this last point, it will be assumed the vast majority of the traffic during the morning peak hour is heading towards the city, while the traffic during the afternoon peak hour is leaving the city.

* All toll booths in each of the plazas is considered to be the same in terms of technology and service time. The only difference that will be considered would be that of automatic and manual toll booths.

* Service time is considered independent of the length of the queue.

#### Estimation of traffic intensity per toll booth
In the previous section we discussed the need of computing the arrival rate $\lambda$ and service rate $\mu$ in order to know the traffic intensity $\rho$. The former can be derived directly from the information provided by the dataset, given that each of its records indicates the amount of vehicles that arrived and went through the toll booths per hour. In order to estimate $\mu$ a sample toll booth service time was taken from the _Alberti_ toll booth plaza. Additionally, only _cars_ volume will be taken into account, as sample consists only on cars measurements.

The next three section will consist on a detailed explanation of each of the three variables of $\rho = \frac{\lambda}{S\mu}$, with the objective of estimating the utilization $\rho$ of the _Alberti_ at each hour of the day.

##### Arrivals
At the moment of writing the most recent information the dataset holds is from January 2019. The rationale for using the most recent information to define the arrival rate is that the volume of vehicles going through toll booths steadily increased through the years, so using information from prior years can lower the average arrival volumes. Figure 4 shows the average amount of vehicles that arrive to the _Alberti_ toll booth per hour.

```{r albertivehiclesperhour, echo=FALSE}
years_to_filter_by <- c(2019) # 2014:2019
# Filtering only for Alberti information, on the years where there is more
# recent information (which is from 2019 onwards), and only for the top 2
# most frequent payment methods.
toll_booth_ff <- unified_ff %>% as_tibble() %>%
  filter (
    year_ %in% years_to_filter_by
    & toll_booth_name == "Alberti"
    & payment_method %in% c("Automatic", "Cash")
    & vehicle_type == "Car"
  )

# We'll need to know the amount of days that our dataset contains, as further
# below it will be used to compute the amount of vehicles per hour, and each
# hour is present in every day.
amount_of_days <- toll_booth_ff %>%
  distinct (
    year_,
    month_,
    day_
  ) %>%
  nrow()

payments <- toll_booth_ff %>%
  group_by (
    payment_method,
    start_hour
  ) %>% 
  summarise (
    # Divide by 4 as each row is repeated 4 times due to the oil dataset join.
    amount = 0.25 * sum(amount) / amount_of_days
  )

# Average hourly payments performed by vehicles that arrive at toll booths.
payments_per_hour <- payments %>%
  group_by (
    start_hour
  ) %>%
  summarise (
    amount = sum(amount)
  )

ggplot(data = payments_per_hour,
       aes(x = substr(start_hour, 1, 2), y = amount, group = 1)) +
  geom_line(linetype = "dashed") +
  geom_point(color = "blue") +
  labs(x = "Hour of day",
       y = "Volume of vehicles",
       title = 
         "Hourly average volume of cars in the Alberti toll booth plaza",
       caption = "Figure 4") +
  theme_minimal()

rm(payments_per_hour, amount_of_days, toll_booth_ff)
```

The previous figure presents two peaks which correspond to the timeframes of 9am to 11am and 6pm to 8pm. It would be expected that the utilization of the system, length of the queues, and the time spent on them increases when approaching a peak hour. Figure 5 ilustrates the breakdown of the previously shown volume per payment method, which clearly indicates the peaks are mostly driven by users leaning towards the use of the automatic payment method.

```{r arrivalrate, echo=FALSE}
# Arrival rate.
lambda_manual <- payments$amount[payments$payment_method == "Cash"]
lambda_automatic <- payments$amount[payments$payment_method == "Automatic"]
lambda <- lambda_automatic + lambda_manual
ggplot(data = payments,
       aes(x = substr(start_hour, 1, 2), y = amount, group = payment_method)) +
  geom_line(aes(color = payment_method)) +
  labs(x = "Hour of day",
       y = "Volume of vehicles",
       title = 
         "Hourly vehicle volume breakdown per payment method",
       caption = "Figure 5") +
  theme_minimal()

rm(lambda_manual, lambda_automatic)
```


##### Service time
The _Alberti_ toll booth plaza has a total of 5 toll booths, 2 of them for accessing the _25 de Mayo_ highway, and other 3 for leaving it. A sample of 70 observations of vehicles being serviced on this toll booth was taken on a Saturday, which yielded the distribution presenteed in Figure 6.

```{r attentionsample, echo=FALSE}
datasets_dir <- "../datasets/sources"
attention <- data.frame (
  read.csv (
    file = file.path(datasets_dir, "attention.csv"), 
    header = T, 
    dec = ".",
    stringsAsFactors = T
  )
)

ggplot(data = attention, aes(x = time, fill = type)) + 
  geom_density(alpha = .3) +
  labs(x = "Service time (s)", 
       y = "Density",
       title = "Density of each payment method",
       caption = "Figure 6") +
  theme_minimal()
```

The density function for the automatic payment method is more narrow than its manual counter part given that the latter could be impacted by many factors like the driver not having exact change prepared to pay, or the toll booth employee taking more or less time to service different clients. The average service time for the automatic payment method is `r mean(attention$time[attention$type == "automatic"])` seconds, while the average service time for manual payments is `r mean(attention$time[attention$type == "manual"])` seconds.

One characteristic of _Alberti_'s toll booths is they are both automatic and manual, so depending on the payment method of choosing of the driver, they can pay with cash or with an electronic tag. In summary, Figure 5 tells us there are moments of the day where automatic payments have more impact, while Figure 6 illustrates the densities of each payment method, and how service time differs between them. Such difference has to be taken into account when computing the service rate $\mu$ for the estimation of the traffic intensity $\rho$.

```{r servicerate, echo=FALSE}
# Observations were measured in seconds. Given that the volume of vehicles
# paying with cash is not the exact same volume as those paying automatically,
# a weighted average of both payments methods is used to compute both the 
# arrival and service rate.
share <- payments %>%
  group_by (
    start_hour
  ) %>%
  summarise (
    manual = sum(amount[payment_method == "Cash"]) / sum(amount),
    automatic = sum(amount[payment_method == "Automatic"]) / sum(amount)
  )
# Service rate. Given that the volume of vehicles paying with cash is not the 
# exact same volume as those paying automatically, a weighted average of 
# payments methods is used to compute the service rate.
t_automatic <- mean(attention$time[attention$type == "automatic"]) / 3600
t_manual <- mean(attention$time[attention$type == "manual"]) / 3600
mu <- 1 / (share$automatic * t_automatic + share$manual * t_manual)

rm(t_automatic, t_manual, share)
```

For this reason the service rate has to be expressed as a weighted sum which is function of the mean service times and its relative hourly weight. Such relation can be expressed as $\mu = \frac{1}{\alpha_it_a + \beta_it_m}$ where $\alpha_i$ indicates the percentage of vehicles paying with the automatic payment method at the hour $i$, and $\beta_i$ indicates the percentage of vehicles paying manually at the hour $i$, or $\beta = 1 - \alpha$. The resulting hourly values of $\mu$ are presented in Figure 7

```{r graph, echo=FALSE}
# Adds the hour of the day to each of the hourly service rate values.
mu_df <- data.frame (
  time = 0:23,
  mu = mu
)

ggplot(data = mu_df, aes(x = time, y = mu)) +
  geom_line(linetype = "dashed") +
  geom_point(color = "blue") +
  labs(x = "Hour of the day", 
       y = "Amount of vehicles",
       title = "Vehicles served per hour per toll booth",
       caption = "Figure 7") +
  theme_minimal()

rm(mu_df)
```

##### Traffic intensity
Traffic intensity is defined as the ratio between the arrival rate $\lambda$ and the service rate $\mu$, so that $\lambda < \mu$. That condition ensures the system is stable. However, it only accounts for the case of a single server, while in this case  we have $S$ toll booths per plaza. This yields a system stability condition of $\lambda < S\mu$, thus defining $\rho$ as $\rho = \frac{\lambda}{S\mu}$. Following the third assumption in the assumptions section, as all toll booths are considered equal, we can express the total capacity of the system by multipying an individual toll booth capacity by the amount of toll booths in a toll booth plaza.

Given the definitions of $\lambda$ and $\mu$ in the prior two sections, the utilization of the _Alberti_ toll booth plaza can be described by Figure 8, which presents the percentage of utilization of the 5 toll booths in such plaza. The moments where the toll booth plaza is mostly used are peak hours, with 10am having a 73.5% utilization, and 6pm with 72% utilization.

```{r realsample, echo=FALSE}
# Using the empirically computed service rate, the utilization is recomputed.
S <- 5
rho <- lambda / (S * mu)

# Adds the hour of the day to each of the hourly service rate values.
rho_df <- data.frame (
  time = 0:23,
  rho = rho
)

ggplot(data = rho_df, aes(x = time, y = 100 * rho)) +
  geom_line(linetype = "dashed") +
  geom_point(color = "blue") +
  labs(x = "Hour of the day", 
       y = "Percentage of toll booth plaza utilization",
       title = "Hourly utilization of Alberti toll booth plaza",
       caption = "Figure 8") +
  theme_minimal()

rm(rho_df)
```

##### Characterization of the M/M/S system
The system will be modeled as an M/M/S queuing model, where the first *M* stands for a Poisson distributed arrival pattern of vehicles. From here is that we derive $\lambda$, the mean of such distribution. The second *M* stands for an exponential distribution of the service rate $\mu$. Lastly, S stands for the number of servers, in this case toll booths per toll booth plaza, with a First In - First Out discipline. Summing everything together, the system will be considered stable if $\lambda < S\mu$, as previously mentioned.

```{r characteristics, echo=FALSE}
# Probability of having no units in the system (idle system).
p0_ <- function(
  S, # Number of servers.
  rho # Utilization of the system.
) {
  n <- 0:(S-1)
  1 / (
    sum((rho)^n / factorial(n)) + (rho^S) / (factorial(S) * (1 - (rho / S)))
  )
}

# Probability of having n units in the system. This means having `n` vehicles
# arriving at the same time, one to each toll booth. 
pn <- function(
  n, # Amount of vehicles in the system.
  S, # Number of servers.
  rho # Utilization of the system.
) {
  if(!n) {
    p0_(S, rho)
  } else if((n - 1) == 0) {
    (rho / n) * p0_(S, rho) 
  } else {
    pn(n - 1, S, rho)
  }
}

# Average number of units in the queue.
Uq <- function (
  S, # Number of servers.
  rho # Utilization of the system.
) {
  p0_(S, rho) * ((rho^S) / factorial(S)) * (rho / S) / (1 - (rho / S))^2
}

# Average number of units in the system.
Us <- function (
  S, # Number of servers.
  rho # Utilization of the system.
) {
  rho + Uq(S, rho)
}

# Average time waiting in a queue.
Wq <- function (
  S, # Number of servers.
  rho, # Utilization of the system.
  lambda # Arrival rate.
) {
  Uq(S, rho) / lambda
}

# Average time waiting in the system.
Ws <- function (
  S, # Number of servers.
  rho, # Utilization of the system.
  lambda, # Arrival rate.
  mu # Service rate.
) {
  Wq(S, rho, lambda) + (1 / mu)
}
```

###### Probability of N vehicles arriving at a given time
Traffic intensity $\rho$ can be used to determine the probability of having no vehicles queuing in the system at a given moment of time, and is written as $P_0 = \frac{1}{\sum_{n = 0}^{S - 1}\frac{\rho^n}{n!}+\frac{\rho^S}{S!(1-\frac{\rho}{S})}}$, where $S$ is the number of toll booths. Given this definition of $P_0$, the analogoues $P_n$, which would be the probability of having $n$ units in the system, is defined as $P_n = \left ( \frac{\rho}{n} \right ) P_{n-1}$. With our assumption of $90\%$ utilization the probability of having no vehicles in the system is `r round(100 * pn(0, S, rho), 2)`, while the probability of having $S$ vehicles arriving at the $S$ toll booths at the exact same time to be served is `r round(100 * pn(S, S, rho), 2)`.

###### Average number of units in queues
Using the probability of having an empty queue $P_0$ we can derive the average queue length. This metric will then be used to estimate the average waiting time in queues. As this are average times, they will take into account those times where there are no vehicles queuing, and times where there are long queues. For the case of $S$ servers, the average queue length is defined as $U_q(S, \rho) = P_0\frac{\rho^{S+1}}{S!S}\left [ \frac{1}{\left (1 - \frac{\rho}{S} \right )^2} \right ]$, and has an approximate value of `r Uq(S, rho)`. This value is actually very small, and could be rounded to $0$, and can be interpreted as arriving vehicles not having to queue for going through a toll booth.

###### Average waiting time for units in queues
The $U_q(S,\rho)$ metric can be reused to define the time spent waiting on queues as $W_q(S,\rho,\lambda) = U_q/\lambda$. The units of this metric are hours, and for this particular case it's value is `r Wq(S, rho, lambda)`, again very close to $0$. This result goes in tandem with the results from the average number of vehicles per queue, as not having to queue behind other vehicles means not spending that time on queues.
